Designing Types for R, Empirically

### 2020-10-15 paper/main.tex
types

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Designing A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we set out to propose a candidate design for a type
language to describe the arguments and return values of R functions.
The goal is not to propose a final design, but rather a
starting point for an iterative design process.

Fig.~\ref{types} presents our type language. An early design choice
was to stay close to the R language and only depart in small and
hopefully non-controversial ways.  Functions types have the form
\FUN{A_1, \dots, A_n}\T where each $\A_i$ argument is either a type \T
or \VARGS, a variable length argument list. The rest of this section
details and motivates our design.

@@

\paragraph{Scalar vs. Vectors} We distinguish between
vectors of length 1, and vectors of any dimension. The base types, \int,
\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
length 1, and thus a scalar value is also a vector.  Vectors are
homogeneous, in that a vector of doubles contains only doubles.

\paragraph{NA values} In R, a value can be \emph{not available}, written
\NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
well as a \dbl \NA, and they are different.  The type system allows one to
distinguish between vectors that may contain {\NA}s (written \NAVEC\int) and
those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
treated as a vector that may {\NA}s.  We do not allow scalar values to be
\NA and the type \raw does not allow {\NA}s.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we set out to propose a candidate design for a type
language to describe the signatures of R functions.  The goal is not
to propose a final design, but rather a starting point for an
iterative process.

Given the pecularities of the language there are a number of design
choices that need to be reviewed and evaluated. It is uncontroversial
to include types that cover the six kinds of primitive vectors,
furthermore environments and the distinguished null types are commonly
used and must be included. Fig.~\ref{types} presents our type language.

@@

\paragraph{Scalars} We distinguish between
vectors of length 1, and vectors of any dimension. The base types, \int,
\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
length 1, and thus a scalar value is also a vector.  Vectors are
homogeneous, in that a vector of doubles contains only doubles.

\paragraph{Missing} 
Each of basic types has its specific \NA. Thus there is an \int \NA as
well as a \dbl \NA, and they are different.  The type system allows
one to distinguish between vectors that may contain {\NA}s (written
\NAVEC\int) and those who are guarantee to be \NA-free (\VEC\int).  A
\NA-free vector can be treated as a vector that may {\NA}s.  We do not
allow scalar values to be \NA and the type \raw does not allow {\NA}s.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$. It is noteworthy that the elements
of a union are not necessarily disjoint. For instance a \NAVEV\int

@@

\paragraph{Functions.}
Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
$\A_i$ argument is either a type \T or \VARGS, a variable length
argument list. Moreover, a single function's signature can be the
disjunction of a number of individual signatures.
